---
title: "DEI ACS PUMS Data Checks - Indicator 30: Utility Expenses"
author: "Kaitlan Wong"
format: html
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidycensus) 
library(tidyverse) 
library(readxl) 
library(writexl) 
library(dplyr)
```

### **Helpful Resources:**

-   See <https://walker-data.com/census-r/analyzing-census-microdata.html> for guidance on using tidycensus to get ACS PUMS

-   ACS PUMS Data Documentation: <https://www.census.gov/programs-surveys/acs/microdata/documentation.html>

-   2021 ACS Data Dictionary: <https://www2.census.gov/programs-surveys/acs/tech_docs/pums/data_dict/PUMS_Data_Dictionary_2021.pdf>

### **Variable Definitions**

-   WATP: Water cost (yearly cost, use ADJHSG to adjust WATP to constant dollars)

-   GASP: Gas cost (monthly cost, use ADJHSG to adjust GASP to constant dollars)

-   FULP: Fuel cost (yearly cost for fuels other than gas and electricity, use ADJHSG to adjust FULP to constant dollars)

-   ELEP: Electricity cost (monthly cost, use ADJHSG to adjust ELEP to constant

    dollars)

Get 2021 ACS PUMS data.

```{r}
# Get ACS data
pums21 <- get_pums(
  variables = c("AGEP", "RAC1P", "HISP", "PUMA", "GASP", "FULP", "ELEP", "WATP", "ADJHSG"),
  state = "TX",
  survey = "acs1",
  year = 2021,
  recode = TRUE,
)

# View data
head(pums21)
```

Filter for Dallas PUMAs and mutate variables

```{r}
# Filter for Dallas PUMAs
data21 <- pums21 %>%
  filter(PUMA %in% c("02304", "02305", "02306", "02307", "02309", "02310", "02311", "02312", "02313", "02314", "02315", "02316", "02319", "01901", "02001")) %>%
  # recode race/ethnicity
  mutate(race_eth = case_when(
    HISP != "01" ~ "Hispanic",
    HISP == "01" & RAC1P == "1" ~ "White, non-Hispanic",
    HISP == "01" & RAC1P == "2" ~ "Black, non-Hispanic",
    HISP == "01" & RAC1P %in% c("3","4","5") ~ "AIAN, non-Hispanic",
    HISP == "01" & RAC1P %in% c("6","7") ~ "Asian or Pacific Islander, non-Hispanic",
    HISP == "01" & RAC1P %in% c("8","9") ~ "Other, non-Hispanic",
    TRUE ~ NA_character_)
  ) %>%
  # Multiply GASP and ELEP by 12 to get annual cost
  mutate(gas = GASP * 12) %>%
  mutate(electric = ELEP * 12) %>%
  # Add gas, electricity, fuel, and water costs to get total annual utility expenses
  mutate(utilities = gas + electric + FULP + WATP)
```

Create table of median annual utility expenses using household weight variable.

```{r}
library(matrixStats)

# Calculate median utility expenses by race/ethnicity of householder

summary21 <- data21 %>%
  group_by(race_eth) %>%
  summarize(
    median_util = weightedMedian(utilities, WGTP)
  )

summary21

# Calculate the weighted median for all households
summary_all21 <- data21 %>%
  summarize(
    median_inc = weightedMedian(utilities, WGTP)
  )

summary_all21
```

Repeat steps for 2019 data.

```{r}
# Get ACS data
pums19 <- get_pums(
  variables = c("AGEP", "RAC1P", "HISP", "PUMA", "GASP", "FULP", "ELEP", "WATP"),
  state = "TX",
  survey = "acs1",
  year = 2019,
  recode = TRUE,
)

# View data
head(pums19)
```

```{r}
# Filter for Dallas PUMAs
data19 <- pums19 %>%
  filter(PUMA %in% c("02304", "02305", "02306", "02307", "02309", "02310", "02311", "02312", "02313", "02314", "02315", "02316", "02319", "01901", "02001")) %>%
  # recode race/ethnicity
  mutate(race_eth = case_when(
    HISP != "01" ~ "Hispanic",
    HISP == "01" & RAC1P == "1" ~ "White, non-Hispanic",
    HISP == "01" & RAC1P == "2" ~ "Black, non-Hispanic",
    HISP == "01" & RAC1P %in% c("3","4","5") ~ "AIAN, non-Hispanic",
    HISP == "01" & RAC1P %in% c("6","7") ~ "Asian or Pacific Islander, non-Hispanic",
    HISP == "01" & RAC1P %in% c("8","9") ~ "Other, non-Hispanic",
    TRUE ~ NA_character_)
  ) %>%
  # Multiply GASP and ELEP by 12 to get annual cost
  mutate(gas = GASP * 12) %>%
  mutate(electric = ELEP * 12) %>%
  # Add gas, electricity, fuel, and water costs to get total annual utility expenses
  mutate(utilities = gas + electric + FULP + WATP)


library(matrixStats)

# Calculate median utility expenses by race/ethnicity of householder

summary19 <- data19 %>%
  group_by(race_eth) %>%
  summarize(
    median_util = weightedMedian(utilities, WGTP)
  )

summary19

# Calculate the weighted median for all households
summary_all19 <- data19 %>%
  summarize(
    median_inc = weightedMedian(utilities, WGTP)
  )

summary_all19
```

```{r}
# Export tables to Excel

write_xlsx(list("2021" = summary21, "2019" = summary19), "30_Utility Expenses.xlsx")
```
