---
title: "DEI ACS PUMS Data Checks - Indicator 30: Utility Expenses - 2016 Final Version"
author: "Kaitlan Wong, Edited by Coda Rayo-Garza"
format: html
editor: visual
---

### **Helpful Resources:**

-   See <https://walker-data.com/census-r/analyzing-census-microdata.html> for guidance on using tidycensus to get ACS PUMS

-   ACS PUMS Data Documentation: <https://www.census.gov/programs-surveys/acs/microdata/documentation.2016.html#list-tab-1370939201>

-   2016 ACS Data Dictionary: <https://www2.census.gov/programs-surveys/acs/tech_docs/pums/data_dict/PUMSDataDict16.pdf>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidycensus) 
library(tidyverse) 
library(readxl) 
library(writexl) 
library(dplyr)
```

```{r}
#Use this code to look up PUMS variables 
view(pums_variables)
```

We will need:

-   WATP: Water cost (yearly cost, use ADJHSG to adjust WATP to constant dollars)

-   GASP: Gas cost (monthly cost, use ADJHSG to adjust GASP to constant dollars)

-   FULP: Fuel cost (yearly cost for fuels other than gas and electricity, use ADJHSG to adjust FULP to constant dollars)

-   ELEP: Electricity cost (monthly cost, use ADJHSG to adjust ELEP to constant

    dollars)

-   HINCP: Household Income in the past 12 months (ADJINC to adjust HINCP to constant dollars)

-   ADJINC: Adjustment factor for income and earnings dollar amounts

-   ADJHSG: Adjustment factor for housing dollar amounts (6 implied decimal places)

```{r}
#Pull PUMS data using tidycensus 
u16 <- get_pums(
  variables = c("AGEP", "RAC1P", "HISP", "PUMA", "GASP", "FULP", "ELEP", "WATP", "ADJINC", "ADJHSG", "HINCP"),
  state = "TX",
  survey = "acs1",
  year = 2016,
  recode = TRUE,
)

# View data
head(u16)
```

```{r}
# check adjustment factor variables to see what format they are in (implied decimals vs. decimal format)

u16$ADJHSG  # this is already in decimal format, so no need to divide by 1e+06 (unlike 2016 pums dataset)

u16$ADJINC  # this is already in decimal format, so no need to divide by 1e+06
```

Filter for Dallas PUMAS

```{r}
pumas16 <- u16 %>%
  filter(PUMA %in% c("2304", "2305", "2306", "2307", "2309", "2310", "2311", "2312", "2313", "2314", "2315", "2316", "2319", "1901", "2001")) %>%
  # recode race/ethnicity
  mutate(race_eth = case_when(
    HISP != "01" ~ "Hispanic",
    HISP == "01" & RAC1P == "1" ~ "White, non-Hispanic",
    HISP == "01" & RAC1P == "2" ~ "Black, non-Hispanic",
    HISP == "01" & RAC1P %in% c("3","4","5") ~ "AIAN, non-Hispanic",
    HISP == "01" & RAC1P %in% c("6","7") ~ "Asian or Pacific Islander, non-Hispanic",
    HISP == "01" & RAC1P %in% c("8","9") ~ "Other, non-Hispanic",
    TRUE ~ NA_character_)) 
```

## Per PUMS, we need to make adjustments to get constant dollars for some of the variables

One thing to note is that making the adjustments takes care of any potential negative values in the utilities vectors, so we don't need to worry about negative values. However, do double check for NA's.

```{r}

# Check for NAs in the dataframe
has_na <- anyNA(pumas16)

if (has_na) {
  # If there are NAs, remove rows with NAs
  pumas16 <- pumas16[complete.cases(pumas16), ]
  print("NAs have been removed.")
} else {
  print("No NAs found in the dataframe.")
}
```

```{r}
# covert variables to numeric
pumas16$ADJHSG <- as.numeric(pumas16$ADJHSG)
pumas16$GASP <- as.numeric(pumas16$GASP)
pumas16$ELEP <- as.numeric(pumas16$ELEP)
pumas16$FULP <- as.numeric(pumas16$FULP)
pumas16$WATP <- as.numeric(pumas16$WATP)

# Make adjustments to get constant dollars 
pumas16 <- pumas16 %>%
  mutate(adj_gasp = GASP * ADJHSG,
         adj_elep = ELEP * ADJHSG,
         adj_fulp = FULP * ADJHSG,
         adj_watp = WATP * ADJHSG )

#Get annual costs for gas and electricity since these are originally monthly 
pumas16<- pumas16 %>% mutate(gas = adj_gasp * 12) %>%
  mutate(elec = adj_elep * 12) 

#yearly costs for water (WATP) and yearly costs for fuel (FULP) don't need to be multiplied
  
  
# Calculate total annual utility expenses after adjustment for each household
pumas16 <- pumas16 %>%
  mutate(total_annual_utility_expenses = gas + elec + adj_fulp + adj_watp)


# Calculate weighted total annual costs by race/ethnicity
total_costs_by_race <- pumas16 %>%
  group_by(race_eth) %>%
  summarize(weighted_total_annual_costs = sum(total_annual_utility_expenses * WGTP, na.rm = TRUE) / sum(WGTP, na.rm = TRUE))

# View the results -- REMINDER, this is total costs 
print(total_costs_by_race)




```

## Now, to Get Median Utility Expenses

```{r}

# Code 1 (I tried several, including using the scales package and that was not successful)

# Calculate the weighted median annual costs by race/ethnicity
weighted_median <- function(x, w) {
  x <- sort(x)
  w <- w[order(x)]
  n <- sum(w)
  cum_w <- cumsum(w)
  threshold <- n / 2
  median_index <- which(cum_w >= threshold)[1]
  median_value <- x[median_index]
  median_value
}

median_costs_by_race <- pumas16 %>%
  group_by(race_eth) %>%
  summarize(median_annual_costs = weighted_median(total_annual_utility_expenses, w = WGTP))

# View the results
print(median_costs_by_race)



```

## Now about HH Income...

Here we do have to worry about negative numbers. So, we make the adjustment using ADJINC and filter out negative incomes.

```{r}
# Double Checking for Lonely PSU's (a "Lonely PSU" refers to a unique situation where a Public Use Microdata Area (PUMA) contains only one Primary Sampling Unit (PSU))
psu_counts <- table(pumas16$PSU)

# Find the PSUs with only one sampled unit (lonely PSUs)
lonely_psus <- psu_counts[psu_counts == 1]

# Print the lonely PSUs, if any
if (length(lonely_psus) > 0) {
  print(paste("Lonely PSUs:", paste(names(lonely_psus), collapse = ", ")))
} else {
  print("No lonely PSUs found.")
}


#all good 
```

### Getting Median Household Income

```{r}
library(dplyr)
library(srvyr)

# covert variables to numeric
pumas16$ADJINC <- as.numeric(pumas16$ADJINC)
pumas16$HINCP <- as.numeric(pumas16$HINCP)

# Remove households with income less than $0
pumas16 <- pumas16 %>% filter(HINCP >=0)


# Conver ADJINC to numeric
pumas16$ADJINC <- as.numeric(pumas16$ADJINC)


# Adjust household income (HINCP) to constant dollars using 'ADJINC'
# Remember, ADJINC is already in decimal format, so no need to divide by 1e+06
pumas16 <- pumas16 %>%
  mutate(adj_hincp = HINCP * ADJINC)

pumas16$adj_hincp <- as.numeric(pumas16$adj_hincp)



# Group the data by 'race_eth' and calculate the median adjusted household income
median_income_by_race <- pumas16 %>%
  group_by(race_eth) %>%
  summarize(median_adj_hincp = median(adj_hincp))

# Print the result
print(median_income_by_race)
```

### Finally, calculate the median percent of income spent on utilities

```{r}

# Calculate the percentage of income spent on utilities
pumas16 <- pumas16 %>%
  mutate(percent_income_utilities = total_annual_utility_expenses / adj_hincp)

# Group the data by 'race_eth' and calculate the median percentage of income spent on utilities
median_percent_income_utilities_by_race <- pumas16 %>%
  group_by(race_eth) %>%
  summarize(median_percent_utilities = weighted_median(percent_income_utilities, w = WGTP))

# Print the result
print(median_percent_income_utilities_by_race)
```

```{r}
# Export table to Excel

write_xlsx(list("med_inc" = median_income_by_race, "med_util" = median_costs_by_race, "percent_inc" = median_percent_income_utilities_by_race), "30_Utility Expenses_2016.xlsx")
```
